---
title: "Practice Simple Linear Rehression"
author: Qiaoyun Ai
Date: 2026-03-26
editor: source
---

## Step 1

#### Setup

```{r}
# hide unnecessary message
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

# packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(broom)
  library(cowplot)
  library(kableExtra)
  library(skimr)
})
```

#### Using the {tidyverse} `read_csv()` function, load the “Street_et_al_2017.csv” dataset as a “tibble” named **d**.

```{r}
# loading data
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/Street_et_al_2017.csv"
d <- read_csv(f, col_names = TRUE)
head(d)
```

#### A quick exploratory data analysis where you generate the five-number summary (median, minimum and maximum and 1st and 3rd quartile values), plus mean and standard deviation, for each quantitative variable.

```{r}
skim(d) |>
    kable() |>
    kable_styling(font_size = 10, full_width = FALSE)
```

## Step 2

#### Plot brain size (**ECV**) as a function of social group size (**Group_size**), longevity (**Longevity**), juvenile period length (**Weaning**), and reproductive lifespan (**Repro_lifespan**).

```{r}
p1 <- ggplot(data = d, aes(x = Group_size, y = ECV)) + geom_point()
p2 <- ggplot(data = d, aes(x = Longevity, y = ECV)) + geom_point()
p3 <- ggplot(data = d, aes(x = Weaning, y = ECV)) + geom_point()
p4 <- ggplot(data = d, aes(x = Repro_lifespan, y = ECV)) + geom_point()
plot_grid(p1, p2, p3, p4, nrow = 2)
```

## Step 3

#### Derive by hand the ordinary least squares regression coefficients  and  for ECV as a function of social group size.

```{r}
# remove missing values
d_mod <- d |> 
  drop_na(ECV, Group_size)

# slope (beta1)
b1 <- cor(d_mod$ECV, d_mod$Group_size) * 
      sd(d_mod$ECV) / sd(d_mod$Group_size)

# intercept (beta0)
b0 <- mean(d_mod$ECV) - b1 * mean(d_mod$Group_size)

b1
b0
```

## Step 4

#### Confirm that you get the same results using the `lm()` function.

```{r}
m <- lm(ECV ~ Group_size, data = d_mod)
results <- m |>
    summary() |>
    tidy()
results
```

## Step 5

#### Repeat the analysis above for three different major radiations of primates - “catarrhines”, “platyrrhines”, and “strepsirhines”) separately. These are stored in the variable **Taxonomic_group**. 

```{r}
# remove missing values
d_mod <- d |>
  drop_na(ECV, Group_size, Taxonomic_group)

platyrrhini <- d_mod |>
    filter(Taxonomic_group == "Platyrrhini")
catarrhini <- d_mod |>
    filter(Taxonomic_group == "Catarrhini")
strepsirhini <- d_mod |>
    filter(Taxonomic_group == "Strepsirhini")

(d_mod |>
  group_by(Taxonomic_group) |>
  do(tidy(lm(ECV ~ Group_size, data = .))))
```

#### Q: Do your regression coefficients differ among groups? How might you determine this?

#### A: The regression coefficients differ among groups. To test if the differences are significant, I can permute the group labels and compare the observed differences in slopes to a null distribution.

## Step 6

#### For your first regression of ECV on social group size, calculate the standard error for the slope coefficient, the 95% CI, and the *p* value associated with this coefficient by hand. Also extract this same information from the results of running the `lm()` function.

```{r}
# define alpha and degrees of freedom for the regression
alpha <- 0.05
n <- nrow(d_mod)
df <- n - 2

# residuals
residuals <- d_mod$ECV - (b0 + b1 * d_mod$Group_size)

# variance
SSE <- sum(residuals^2)
MSE <- SSE / df

# variation in x
SSX <- sum((d_mod$Group_size - mean(d_mod$Group_size))^2)

# SE
SE_b1 <- sqrt(MSE / SSX)

# 95% CI
CI_b1 <- b1 + c(-1, 1) * qt(0.975, df = df) * SE_b1

# p-value
t_b1 <- b1 / SE_b1
p_b1 <- 2 * pt(abs(t_b1), df = df, lower.tail = FALSE)

SE_b1
CI_b1
p_b1

# extract the same information  using lm() function
lm_results <- tidy(m, conf.int = TRUE)
lm_SE_b1 <- lm_results |> filter(term == "Group_size") |> pull(std.error)
lm_CI_b1 <- lm_results |> filter(term == "Group_size") |> select(conf.low, conf.high)
lm_p_b1  <- lm_results |> filter(term == "Group_size") |> pull(p.value)

lm_SE_b1
lm_CI_b1
lm_p_b1
```

## Step 7

#### Use a permutation approach with 1000 permutations to generate a null sampling distribution for the **slope coefficient**.  You can use either the quantile method (i.e., using quantiles from the actual permutation-based null sampling distribution) or a theory-based method (i.e., using the standard deviation of the permutation-based null sampling distribution as the estimate of the standard error, along with a normal or t distribution), or both, to calculate this p value.

```{r}
set.seed(1)
nperm <- 1000
perm <- numeric(nperm)
for (i in 1:nperm) {
  perm.sample <- d_mod
  perm.sample$Group_size <- sample(perm.sample$Group_size)
  perm[i] <- lm(ECV ~ Group_size, data = perm.sample) |>
    tidy() |>
    filter(term == "Group_size") |>
    pull(estimate)
}

# plot
hist(perm,
     main = "Permutation distribution of slope",
     xlab = "Slope",
     breaks = 30)

abline(v = b1, col = "black", lwd = 3)

# permutation SE
perm_se <- sd(perm)

# permutation p-value
p_perm <- (1 + sum(abs(perm) >= abs(b1))) / (nperm + 1)

# theory-based p-value
p_perm_theory <- 2 * pt(abs(b1) / perm_se, df = df, lower.tail = FALSE)

p_perm
p_perm_theory
```

#### Q: What is it that you need to permute? What is the p value associated with your original slope coefficient?

#### A: I permute the predictor variable (Group_size) relative to the response (ECV) to break their association. The permutation p-value is approximately 0 (p \< 0.001), indicating the slope is highly significant.

## Step 8

#### Use bootstrapping to generate a 95% CI for your estimate of the slope coefficient using both the quantile method and the theory-based method (i.e., using the standard deviation of the bootstrapped sampling distribution as an estimate of the standard error). 

```{r}
set.seed(1)
nboot <- 1000
boot <- numeric(nboot)
for (i in 1:nboot) {
  boot.sample <- d_mod |> sample_n(nrow(d_mod), replace = TRUE)
  boot[i] <- lm(ECV ~ Group_size, data = boot.sample) |>
    tidy() |>
    filter(term == "Group_size") |>
    pull(estimate)
}

# parameters
alpha <- 0.05
p.lower <- alpha / 2
p.upper <- 1 - alpha / 2

# plot bootstrap distribution
hist(boot,
     main = "Bootstrap distribution of slope",
     xlab = "Slope",
     breaks = 30)

abline(v = b1, col = "black", lwd = 3)

# quantile CI
boot_ci_quantile <- quantile(boot, c(0.025, 0.975))

# theory CI
boot_se <- sd(boot)
boot_ci_theory <- b1 + c(-1, 1) * 1.96 * boot_se

boot_ci_quantile
boot_ci_theory

# plot
ggplot(data.frame(stat = boot), aes(x = stat)) +
  geom_histogram(bins = 30, fill = "gray30", color = "white") +
  
  geom_vline(xintercept = boot_ci_quantile, color = "red", size = 1.5) +
  
  geom_vline(xintercept = boot_ci_theory, color = "blue", size = 1.5) +
  
  geom_vline(xintercept = b1, color = "black", size = 1.2) +
  
  labs(
    title = "Bootstrap distribution of slope",
    x = "Slope (beta1)",
    y = "Count"
  ) +
  theme_minimal()
```

#### Q: Do these CIs suggest that your slope coefficient is different from zero?

#### A: Both bootstrap confidence intervals do not include zero, indicating the slope is significantly different from zero.
